Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-03 analysis/paper/paper.qmd
editing the text

@@

Based on typological study, we identify 13 rounded faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers. \[BM: feels like this belongs in results section\]

@@

@@

@@

### degree of reduction

<u>Scar Number</u> The total number of flake scars visible on the spheroid surface.

<u>Removal Ratio</u> Calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area. We used the Removal Ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.

<u>Scar Density Index (SDI)</u> Defined as the number of flake scars divided by the total 3D surface area, the Scar Density Index (SDI) serves as a robust indicator of core reduction intensity, following Clarkson (2013). It has been demonstrated to effectively capture the dynamic nature of the reduction process.

In all three metrics, higher values represent more advanced stages of reduction. Scar counts and surface area measurements were extracted using Geomagic Wrap.

### shape of spheroids

<u>Mean Edge Angle</u> Higher average edge angles suggest a more spherical geometry. Ten edge intersections were randomly sampled per spheroid. Angles were calculated by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing angles using trigonometric functions in R.

<u>Mean Surface Curvature</u> A proxy for surface roughness, and higher mean surface curvature leads to a smoother surface. We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex. The deviation between actual surface points and this fitted plane reflects surface roughness. The calculation steps are completed in Python.

<u>Sphericity</u> Calculated from surface area and volume. Values closer to 1 indicate a shape approaching a geometrically perfect sphere.

<u>Spherical Harmonic Energy </u> Used to quantify shape sysmmtry and regularity, especially at fine scales. We computed the total energy at degree 20, a resolution high enough to capture surface details. Lower energy values indicate smoother and more regular surfaces, while higher values reflect more irregular or angular geometries. Unlike sphericity, which may overestimate sphericality in irregular objects, spherical harmonic energy better captures fine-scale shape deviations.

Together, these variables allow us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies. We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes. This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution. A total of 13 spheroids and sub-spheroids from the site were included in the analysis.

Ideally, all artifacts involved in the spheroid shaping sequence should be considered. However, due to the difficulty in identifying which cores at the site were part of this specific shaping sequence, only the final-stage spheroids were analyzed. This limitation may have influenced the outcomes of the correlation analysis.

@@

We applied spherical harmonics analysis, sphericity index, and centroid offset measurements to both the spheroids and all other multifacial and polyhedron cores in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.

@@

# 读取 Excel 文件
file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data <- read_excel(file_path)

>>>>

...Qianshangying (QSY) ....

@@

```{r}
# 读取 Excel 文件
library(tidyverse)
library(here)
library(readxl)

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data <- read_excel(file_path)

data_clean <- data %>% 
  drop_na()

n_of_artefacts <- nrow(data_clean)
```

Based on typological study, we identify `r n_of_artefacts` rounded faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers. \[BM: feels like this belongs in results section\]

@@

We computed the total energy at degree 20, a resolution high enough to capture surface details. Lower energy values indicate smoother and more regular surfaces, while higher values reflect more irregular or angular geometries. Unlike sphericity, which may overestimate sphericality in irregular objects, spherical harmonic energy better captures fine-scale shape deviations.

@@

## Efficiency of shaping strategy

@@

### Degree of reduction

To compute the degree of reduction we collected data on three metrics: scar number, removal ratio and scar density index (SDI). We measured scar number as total number of flake scars visible on the spheroid surface. Removal ratio was calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area. \[BM: add a sentence to define cortex ratio and give citation.\] We used the removal ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators. SDI is defined as the number of flake scars divided by the total 3D surface area, the SDI serves as a robust indicator of core reduction intensity, following @clarksonMeasuringCoreReduction2013. It has been demonstrated to effectively capture the dynamic nature of the reduction process.In all three metrics, higher values represent more advanced stages of reduction. Scar counts and surface area measurements were extracted using Geomagic Wrap.

### Spheroid shape

We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy, along with mean edge angle, mean surface curvature. Higher average edge angles suggest a more spherical geometry. Ten edge intersections were randomly sampled per spheroid. Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R. Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface. We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex [@Muller2023]. The deviation between actual surface points and this fitted plane reflects surface roughness. The calculation steps were completed in Python.

Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies. We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes. This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution. A total of `r n_of_artefacts` final-stage spheroids and sub-spheroids from the site were included in the analysis.

@@

We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = ) in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.

@@
